---
title: 'Practical 2: Electronic Medical Records and NLP'
author: "Yao Li"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  cache      = TRUE,
  fig.width  = 9,
  fig.height = 5
)

set.seed(42)  # global seed for reproducibility

library(here)        # reproducible file paths
library(tidyverse)   # dplyr, ggplot2, stringr, tidyr, readr, forcats
library(tidytext)    # tidy text mining
library(SnowballC)   # Porter stemmer
library(textstem)    # lemmatization
library(clinspacy)   # MTSamples dataset
library(topicmodels) # Latent Dirichlet Allocation
library(scales)      # axis label helpers
library(sessioninfo) # session reproducibility report
library(reshape2)    # data reshaping library
```

If the above code-block errors, install any missing libraries
(`install.packages(c('LIBRARYNAME1', 'LIBRARYNAME2'))`)

## Introduction

This practical explores natural language processing (NLP) of
unstructured clinical free-text from electronic medical records (EMR).
You will work through an analysis going from raw text to an unsupervised
topic model using the `MTSamples` medical transcription dataset.

The analyses you will do mirror some of those taken in the assigned
readings and refer back to them for some questions:

- **Sun et al. (2024)** (Latent Dirichlet Allocation (LDA)-based topic
  modelling for clinical social work notes)
- **Chung et al. (2024)** (GPT-4 Turbo for classifying perioperative
  outcomes from pre-operative notes).

**Learning objectives:** by the end of this practical you should
understand how to:

- apply regular expressions to search clinical free-text
- tokenise, normalise, stem, and lemmatize medical text
- compute TF-IDF to identify specialty-discriminating terms
- build a document-term matrix and fit an LDA topic model on clinical
  notes
- compare traditional NLP with LLM-based approaches

------------------------------------------------------------------------

## Data Loading and Exploration

This lab is going to use the `MTSamples` medical transcription dataset.
You can find the original source of this data
[here](https://www.mtsamples.com/) (ignore any HTTP vs HTTPS related
warnings).

```{r load-data}
raw_data <- dataset_mtsamples()
glimpse(raw_data)
```

I've not supplied a data dictionary with this dataset. This is a VERY
common situation with real EMR data.

**Q1a (2 marks):** Based on the `glimpse()` output create a simple data
dictionary (variable name with a 1-2 sentence explanation of the
variable).

note_id: A unique integer identifier assigned to each clinical note.

description: A short text summary describing the patient's presentation,
symptoms, or reason for the medical encounter.

medical_specialty: The medical specialty associated with the clinical
note.

sample_name: A descriptive name or title for the specific case or
procedure represented by the note.

transcription: The full transcribed text of the clinical report or
encounter.

keywords: A collection of keywords or terms related to the clinical
note.

**Q1b (2 marks):** What sort of information are you missing about how
this dataset was created that could be important to your analyses?

Demographic information. There are a lot of records which does not have
age, gender, etc.

**Q1c (2 marks):** How many distinct medical specialties are
represented? What are the most common 5? Create a ggplot bar graph
showing how many transcripts come from each specialty.

40, the most common 5 are Surgery, Consult - History and Phy,
Cardiovascular / Pulmonary, Orthopedic and Radiology.

```{r specialty-bar}
#dplyr and ggplot code to count how many medical specialties are represented in the dataset and create a bar graph of them

library(dplyr)
library(ggplot2)

n_specialties <- raw_data |>
  summarise(n = n_distinct(medical_specialty))

n_specialties

top5_specialties <- raw_data |>
  count(medical_specialty, sort = TRUE) |>
  slice_head(n = 5)

top5_specialties

specialty_counts <- raw_data |>
  count(medical_specialty, sort = TRUE)

ggplot(specialty_counts,
       aes(x = reorder(medical_specialty, n), y = n)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Number of Transcripts by Medical Specialty",
    x = "Medical Specialty",
    y = "Number of Transcripts"
  ) +
  theme_minimal()
```

To make patterns slightly easier to interpret for the sake of learning,
we will simplify our dataset down to just transcripts from three
specialties that have relatively distinct clinical vocabularies:
neurological medicine, musculoskeletal/orthopaedic, and diagnostic
imaging. In real analyses this is something you would have to infer from
the data as part of your exploratory data analysis.

```{r filter-data}
filtered_data <- raw_data |>
  filter(medical_specialty %in% c("Orthopedic", "Radiology", "Neurology"))

count(filtered_data, medical_specialty)
```

------------------------------------------------------------------------

## Regular Expressions

Now we are going to learn how to use the `stringr` library for regular
expressions in R. Regular expressions (regex) enable flexible
pattern-based searches over text. As discussed in the lecture, they are
widely used in NLP (including as part of the framework around LLM
systems) for finding measurements, medications, and semi- structured
values embedded in free-text.

Key `stringr` functions:

| Function                  | Purpose                           |
|---------------------------|-----------------------------------|
| `str_detect(x, pat)`      | Returns `TRUE`/`FALSE` per string |
| `str_count(x, pat)`       | Number of non-overlapping matches |
| `str_extract(x, pat)`     | First match                       |
| `str_extract_all(x, pat)` | All matches (returns a list)      |

Key regex special characters (in R strings, backslashes must be doubled
— e.g., `\d` is written `\\d`):

| Pattern | Meaning | Example match |
|------------------------|------------------------|------------------------|
| `\\d` | Any digit `[0-9]` | `"120"` in `"120 mg"` |
| `\\D` | Any non-digit | `"mg"` in `"120 mg"` |
| `\\w` | Word character `[a-zA-Z0-9_]` | `"bp"` in `"bp_value"` |
| `\\W` | Non-word character | `" "` in `"120 mg"` |
| `\\s` | Whitespace (space, tab, newline) | `" "` in `"120 mg"` |
| `\\S` | Non-whitespace | `"120"` in `"120 mg"` |
| `\\b` | Word boundary | matches position between `"120"` and `" "` |
| `.` | Any character except newline | `"1"`, `"a"`, `"/"` |
| `^` | Start of string | `"BP"` in `"BP: 120/80"` |
| `$` | End of string | `"mmHg"` at end of `"120/80 mmHg"` |
| `*` | 0 or more of preceding | `"\\d*"` matches `""` or `"123"` |
| `+` | 1 or more of preceding | `"\\d+"` matches `"1"` or `"123"` |
| `?` | 0 or 1 of preceding (optional) | `"mmHg?"` matches `"mmH"` or `"mmHg"` |
| `{n}` | Exactly *n* of preceding | `"\\d{3}"` matches `"120"` |
| `{n,m}` | Between *n* and *m* of preceding | `"\\d{2,3}"` matches `"12"` or `"120"` |
| `[abc]` | Any one character in set | `"[aeiou]"` matches any vowel |
| `[^abc]` | Any character *not* in set | `"[^0-9]"` matches non-digits |
| `(a\|b)` | Either `a` or `b` | `"(mg\|mcg)"` matches `"mg"` or `"mcg"` |
| `( )` | Capture group | `"(\\d+) mg"` captures the number |
| `[:alpha:]` | Any letter (POSIX class) | equivalent to `[a-zA-Z]` |
| `[:digit:]` | Any digit (POSIX class) | equivalent to `\\d` |
| `[:alnum:]` | Any letter or digit | equivalent to `[a-zA-Z0-9]` |
| `[:space:]` | Any whitespace (POSIX class) | equivalent to `\\s` |
| `[:punct:]` | Any punctuation character | `"."`, `","`, `"!"` |

### Finding blood pressure measurements

Blood pressure is typically recorded as systolic/diastolic (e.g.,
"120/80" or "130/85 mmHg"). We can therefore look for a pair of 2-3
digit numbers (`\\d{2,3}`) separated by a `/` (i.e.,
`\\d{2,3}/\\d{2,3}`) optionally followed by a pressure-related `mmHg`
unit (with or without a preceding space `\\s*`). The special regex
characters like `\\d` and `\\s` have that extra `\` so they are
"escaped" and not treated as a literal character.

```{r regex-bp}
bp_pattern <- "\\d{2,3}/\\d{2,3}(\\s*mmHg)?"


filtered_data |>
  mutate(has_bp = str_detect(transcription, bp_pattern)) |>
  count(medical_specialty, has_bp) |>
  pivot_wider(names_from = has_bp, values_from = n,
              names_prefix = "bp_") |>
  mutate(pct_with_bp = percent(bp_TRUE / (bp_TRUE + bp_FALSE),
                                       accuracy = 0.1))
```

### Counting medication dosages

Now let's look for medication doses. We can make a regex that finds one
of the standard dosage units (e.g., `mg` `mcg`, `ml`, `mL` `unit`)
preceded by some sort of number (`\\d+`) with or without a decimal point
(`\\.?\\d*`).

```{r regex-dose}
dose_pattern <- "\\d+\\.?\\d*\\s*(mg|mcg|ml|mL|units?)"

filtered_data |>
    mutate(n_doses = str_count(transcription, dose_pattern)) |>
    filter(n_doses > 0) |>
    group_by(medical_specialty) |>
    summarise(mean_doses   = round(mean(n_doses), 1),
              median_doses = median(n_doses))
```

**Q2 (3 marks):** Write a regular expression to find temperature
measurements in clinical text (e.g., "98.6 F", "37.2°C", "temperature of
38.1", "afebrile"). Report how many notes in each specialty contain at
least one temperature mention.

Regex builders like [this
website](https://spannbaueradam.shinyapps.io/r_regex_tester/) can make
it a lot quicker to iterate and improve your regular expressions!

55 in neurology, 42 in radiology, 18 in orthopedic.

```{r q2}
# Your code here

temp_regex <- "(?i)\\b(?:temperature\\s+of\\s+\\d+(?:\\.\\d+)?|\\d+(?:\\.\\d+)?\\s*°?\\s*[FC]|afebrile)\\b"


temp_by_specialty <- filtered_data |>
  mutate(
    temp_mentioned = str_detect(transcription, temp_regex)
  ) |>
  group_by(medical_specialty) |>
  summarise(
    notes_with_temp = sum(temp_mentioned),
  ) |>
  arrange(desc(notes_with_temp))

temp_by_specialty

```

------------------------------------------------------------------------

## Text Normalisation

A key stage in traditional text analysis is to normalise raw text —
reducing variability due to capitalisation, punctuation, and common
function words — before any modelling or analysis.

### Tokenisation and stop word removal

`tidytext::unnest_tokens()` splits text into individual tokens and
lowercases everything. We then strip residual punctuation, drop very
short tokens, and remove stop words (common words like "the", "of",
"and" that carry little domain information):

```{r tokenise}
tokenised_data <- filtered_data |>
  unnest_tokens(word, transcription, to_lower = TRUE) |>
  mutate(word = str_replace_all(word, "[^[:alpha:]]", "")) |>
  filter(str_length(word) > 1) |>
  anti_join(stop_words, by = "word")

glimpse(tokenised_data)
```

**Q3 (2 marks):** How many stop words are in the `stop_words` list, how
many different `lexicon` are included, and which `lexicon` contains the
most stop words? Answer these questions using code below and create a
bargraph of the number of unique stop words per lexicon.

1149 stop words. 3 different lexcion. SMART contains the most stop
words.

```{r stop-word-lexicons}
nrow(stop_words)

n_distinct(stop_words$lexicon)

lexicon_counts <- stop_words |>
  group_by(lexicon) |>
  summarise(
    n_stop_words = n(),
    n_unique_words = n_distinct(word)
  ) |>
  arrange(desc(n_stop_words))


ggplot(lexicon_counts,
       aes(x = reorder(lexicon, n_unique_words),
           y = n_unique_words)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Number of Unique Stop Words per Lexicon",
    x = "Lexicon",
    y = "Unique Stop Words"
  ) +
  theme_minimal()
```

**Q4 (2 marks):** How many *unique* unigrams (distinct words) are there
in total across all the cleaned and tokenised transcriptions for each
specialty? Which specialty has the largest vocabulary and which the
smallest? Complete the code below, then provide a 1-2 sentence
interpretation of the output.

Neurology has the largest vocabulary and Radiology has the smallest.
Neurology has higher mean number of unique words which means it has
longer or more diverse notes whereas radiology shorter or more
repetitive notes.

```{r q4}
# Uncomment and complete this pipeline-fill in the blank inside n_distinct()
tokenised_data |>
  group_by(medical_specialty) |>
  summarise(unique_words = n_distinct(word))

# Now calculate the mean number of unique unigrams per NOTE for each specialty
  tokenised_data |>
    group_by(medical_specialty, note_id) |>
    summarise(unique_words = n_distinct(word), .groups = "drop") |>
    group_by(medical_specialty) |>
    summarise(mean_unique_words = mean(unique_words))

```

### Bigrams

We can also combined consecutive words into two-word phrases (bigrams)
which better capture medical compound terms that unigrams split apart
(e.g., "blood pressure", "right knee", "general anaesthesia"):

```{r bigrams}
bigrams <- filtered_data |>
  unnest_tokens(bigram, transcription, token = "ngrams",
                n = 2, to_lower = TRUE) |>
  filter(!is.na(bigram)) |>
  mutate(word1 = word(bigram, 1),
         word2 = word(bigram, 2)) |>
  filter(!word1 %in% stop_words$word,
         !word2 %in% stop_words$word,
         !str_detect(word1, "\\d"),
         !str_detect(word2, "\\d")) |>
  select(-word1, -word2)

bigrams |>
  count(medical_specialty, bigram, sort = TRUE) |>
  slice_max(n, n = 6, by = medical_specialty, with_ties = FALSE) |>
  arrange(medical_specialty)
```

**Q5 (1 mark):** What is the most common bigram in each specialty?

lower extremities for neurology, preoperative diagnosis for orthopedic,
carotid artery for radiology.

```{r bigram_totals}
bigrams |>
  count(medical_specialty, bigram, sort = TRUE) |>
  group_by(medical_specialty) |>
  slice_max(n, n = 1, with_ties = FALSE)

```

### Sentence segmentation

Another common text processing step is to split our text into distinct
sentences. Again our old friend `unnest_tokens` can do a lot of the
heavy lifting!

```{r sentences}
sentences_data <- filtered_data |>
  unnest_tokens(sentence, transcription,
                token = "sentences", to_lower = FALSE)

sentences_data |>
  group_by(medical_specialty) |>
  summarise(total_sentences  = n(),
            n_notes          = n_distinct(note_id),
            mean_per_note    = round(total_sentences / n_notes, 1))
```

**Q6 (1 mark):** How many *unique* sentences are there in each
specialty? Uncomment and complete the code below (replace each `___`
with the correct column name):

```{r q6}
sentences_data |>
  group_by(medical_specialty) |>
  summarise(
    unique_sentences = n_distinct(sentence))
```

## Stemming and Lemmatization

However, given how noisy language is and the many variants of the same
word (alternative suffixes) like pluralisation or possessive
apostrophes. We also often want to "normalise" words to a common base
form. There are two main approaches to this:

- **Stemming**-crude suffix stripping, fast but imprecise: "operating"
  to "operat" (`SnowballC::wordStem`)

- **Lemmatization**-linguistically informed root form: "operating" to
  "operate" (`textstem::lemmatize_words`)

  We can apply these normalisations using the associated library
  functions.

```{r stem-vs-lemma}
word_sample <- tokenised_data |>
  distinct(word) |>
  mutate(stem  = SnowballC::wordStem(word, language = "english"),
         lemma = textstem::lemmatize_words(word))

word_sample |>
  filter(word != stem | word != lemma) |>
  select(word, stem, lemma) |>
  slice_head(n = 15)
```

**Q7 (1 mark):** Does lemmatising or stemming result in fewer unique
words?

stemming results in fewer unique words.

```{r stemming_v_lemmatising}
word_sample |>
  summarise(
    original_vocab = n(),
    stem_vocab     = n_distinct(stem),
    lemma_vocab    = n_distinct(lemma)
  )
```

**Q8 (2 marks):** In 2-3 sentences explain why a general-purpose English
lemmatizer is likely to perform poorly on medical transcription text.
Give at least two examples of medical terms that would likely be handled
incorrectly and describe what goes wrong.

Because it is trained on everyday vocabulary and grammatical patterns,
not specialised clinical terminology, abbreviations, or domain-specific
morphology. As a result, it often fails to recognise medical words as
valid lemmas and either leaves them unchanged or incorrectly maps them
to unrelated base forms. For example, a term like “cardiomegaly”
(enlarged heart) may be left unchanged even though medically it encodes
a compositional concept, while “hypertension” might be incorrectly split
or not reduced meaningfully, losing its relationship to variants like
“hypertensive.”

### Applying lemmatisation to the full dataset

Now let's lemmatise the full dataset:

```{r lemmatize, cache=TRUE}
unique_lemmas <- tokenised_data |>
  distinct(word) |>
  mutate(lemma = textstem::lemmatize_words(word))

lemmatised_data <- tokenised_data |>
  left_join(unique_lemmas, by = "word")
```

We can now look more easily at the most frequent (lemmatised) words per
specialty

```{r top-lemmas-plot}
lemmatised_data |>
  count(medical_specialty, lemma, sort = TRUE) |>
  slice_max(n, n = 8, by = medical_specialty, with_ties = FALSE) |>
  ggplot(aes(n, reorder_within(lemma, n, medical_specialty),
             fill = medical_specialty)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~medical_specialty, scales = "free_y") +
  scale_y_reordered() +
  labs(title = "Most frequent lemmas per specialty",
       x = "Count", y = NULL)
```

## TF-IDF: Finding Discriminating Terms

Unfortunately, as you can see above, raw counts are dominated by generic
medical words ("patient", "history", "left"). TF-IDF (term
frequency–inverse document frequency) down-weights terms common across
all documents and up-weights terms characteristic of a specific
category. Sun et al. (2024) used word-frequency analysis as a first step
before LDA to explore the vocabulary of their social work notes.

```{r tfidf}
specialty_tfidf <- lemmatised_data |>
  count(medical_specialty, lemma) |>
  bind_tf_idf(lemma, medical_specialty, n)

specialty_tfidf |>
  group_by(medical_specialty) |>
  slice_max(tf_idf, n = 25, with_ties = FALSE) |>
  ggplot(aes(tf_idf,
             reorder_within(lemma, tf_idf, medical_specialty),
             fill = medical_specialty)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~medical_specialty, scales = "free") +
  scale_y_reordered() +
  labs(title = "Top 25 lemmas by TF-IDF per specialty",
       x = "TF-IDF", y = NULL)
```

**Q9 (2 marks):** Do the top TF-IDF terms make clinical sense for each
specialty? Identify one term that surprises you from any specialty.
Complete the code below by replacing `"your_term_here"` with that term
to pull an example transcription containing it, then briefly explain its
clinical context.

Not necessarily. This example shows a common limitation of TF-IDF in
clinical text: it can highlight documentation language (e.g., “word”)
rather than true clinical concepts. The case describes a 31-year-old
woman with acute neurological symptoms, including Confusion and altered
mental status, slurred speech (dysarthria) Right facial weakness and
right-sided motor drift Emotional lability and behavioral changes
Possible focal neurological deficits on imaging (suggesting ischemic
events in basal ganglia/caudate) This presentation is highly suggestive
of a neurological disorder affecting language and motor control, such
as: Acute ischemic stroke (supported by MRI findings) Possible
encephalitic or inflammatory process (given prior “flu-like illness” and
behavioral changes)

```{r q9}
filtered_data |>
  filter(str_detect(transcription, "word")) |>
  slice(1) |>
  pull(transcription) |>
  str_trunc(600)
```

### Comparing lemma proportions across specialties

Now let's compare across specialties. This plot shows each lemma's
proportion in Orthopedic notes (y-axis) against its proportion in
Radiology or Neurology (x-axis, one facet each). Terms near the dashed
diagonal are used similarly in both specialties; terms far from it are
specialty-specific.

```{r proportion-plot}
specialty_props <- lemmatised_data |>
  count(medical_specialty, lemma) |>
  group_by(medical_specialty) |>
  mutate(proportion = n / sum(n)) |>
  ungroup() |>
  pivot_wider(id_cols     = lemma,
              names_from  = medical_specialty,
              values_from = proportion,
              values_fill = 0) |>
  pivot_longer(c(Radiology, Neurology),
               names_to  = "comparison_specialty",
               values_to = "proportion")

ggplot(specialty_props |> filter(proportion > 0, Orthopedic > 0),
       aes(x = proportion, y = Orthopedic,
           colour = abs(Orthopedic - proportion))) +
  geom_abline(colour = "grey40", linetype = 2) +
  geom_jitter(alpha = 0.1, size = 1.8, width = 0.0003, height = 0.0003) +
  geom_text(aes(label = lemma), check_overlap = TRUE,
            vjust = 1.5, size = 2.5) +
  scale_x_log10(labels = percent_format(accuracy = 0.001)) +
  scale_y_log10(labels = percent_format(accuracy = 0.001)) +
  scale_colour_gradient(limits = c(0, 0.003),
                        low = "darkslategray4", high = "gray75") +
  facet_wrap(~comparison_specialty) +
  theme(legend.position = "none") +
  labs(y = "Orthopedic", x = "Comparison specialty",
       title = "Lemma proportions: Orthopedic vs. Radiology and Neurology")
```

**Q10 (2 marks):** What does this plot tell you about the lexical
similarity between Orthopedic and Neurology compared with Orthopedic and
Radiology? Is this what you would expect given the clinical roles of
each specialty? Justify your answer in 2-4 sentences.

Orthopedic vocabulary is not strongly differentiated between Neurology
and Radiology in this dataset. In other words, Orthopedic notes appear
to share a similar proportion of terms with both specialties, meaning
the lexical overlap is broadly comparable in both comparisons rather
than one showing clearly stronger similarity. Clinically, this suggests
that Orthopedics draws on a fairly general anatomical and procedural
vocabulary that overlaps with both Radiology (imaging, structure,
fractures) and Neurology (anatomy, motor function, injury effects),
rather than being uniquely closer to one specialty.

------------------------------------------------------------------------

## Topic Modelling with Latent Dirichlet Allocation

Latent Dirichlet Allocation (LDA) is an unsupervised probabilistic model
that treats each document as a mixture of topics and each topic as a
probability distribution over words. Unlike supervised classifiers, LDA
discovers thematic structure from the text alone-no specialty labels are
used during training. This mirrors the approach of Sun et al. (2024),
who applied LDA to clinical social work notes to uncover latent SDoH
themes without predefined categories.

Two key outputs of a fitted LDA model:

| Output | Symbol | Interpretation |
|------------------------|------------------------|------------------------|
| Word–topic matrix | β (beta) | Probability that word *w* is generated by topic *k* |
| Document–topic matrix | γ (gamma) | Proportion of document i.e., transcript *d* attributed to topic *k* |

### Building the document-term matrix

We aggregate lemma counts per note and apply two filters before building
the DTM:

1.  **Minimum document frequency** - drop terms appearing in fewer than
    5 notes to reduce sparsity.
2.  **TF-IDF filtering** - drop terms whose mean TF-IDF (computed per
    note) falls in the bottom quartile. Words that are common across
    nearly all notes have low IDF and therefore low TF-IDF; retaining
    them would give LDA noisy, non-discriminative signal.

```{r dtm-build, cache=TRUE}
term_counts <- lemmatised_data |>
  count(note_id, lemma)

# Keep only terms present in >= 5 notes
doc_freq_filter <- term_counts |>
  distinct(note_id, lemma) |>
  count(lemma, name = "doc_count") |>
  filter(doc_count >= 5)

# Keep only terms whose mean per-note TF-IDF is above the 25th percentile
tfidf_filter <- term_counts |>
  bind_tf_idf(lemma, note_id, n) |>
  group_by(lemma) |>
  summarise(mean_tfidf = mean(tf_idf), .groups = "drop") |>
  filter(mean_tfidf > quantile(mean_tfidf, 0.25))

emr_dtm <- term_counts |>
  semi_join(doc_freq_filter, by = "lemma") |>
  semi_join(tfidf_filter,    by = "lemma") |>
  cast_dtm(note_id, lemma, n)

emr_dtm <- emr_dtm[rowSums(as.matrix(emr_dtm)) > 0, ]

cat("DTM:", nrow(emr_dtm), "documents ×", ncol(emr_dtm), "terms\n")
```

### Fitting the LDA model

We fit LDA with `k = 3` topics-one per specialty-so we can directly
compare the discovered topics against the known specialty labels. Again
this would likely be something you explore or optimise using more
sophisticated hierarchical models.

```{r fit-lda, cache=TRUE}
set.seed(42)
k_topics <- 3
emr_lda  <- topicmodels::LDA(emr_dtm, k = k_topics,
                              control = list(seed = 42))
```

### Word–topic probabilities (β)

The β matrix tells us which words are most characteristic of each topic:

```{r lda-beta-plot}
lda_beta <- tidy(emr_lda, matrix = "beta")

lda_beta |>
  group_by(topic) |>
  slice_max(beta, n = 12, with_ties = FALSE) |>
  ggplot(aes(beta,
             reorder_within(term, beta, topic),
             fill = factor(topic))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~topic, scales = "free_y",
             labeller = labeller(topic = function(x) paste("Topic", x))) +
  scale_y_reordered() +
  labs(title = "Top 12 words per LDA topic (β probabilities)",
       x = "β  (word–topic probability)", y = NULL)
```

**Q11 (2 marks):** Based on the top words in the β plot, give each topic
a brief descriptive clinical label (e.g., "imaging", "musculoskeletal",
"neurological"). Which topic do you think corresponds most closely to
which specialty, and why? Does the vocabulary in each topic match your
expectations from the TF-IDF results in Section 6?

Topic 1: neuroimaging, it corresponds most closely to neurology. Because
it is most consistent with neurological imaging and brain pathology
descriptions and has a strong emphasis on brain lesions, arteries, and
tumor evaluation.

Topic 2: spine, it corresponds most closely to radiology. It is highly
imaging-centric vocabulary involving spinal structures and likely
reflects radiology reports interpreting MRI/CT of the spine.

Topic 3: Musculoskeletal injury, it corresponds most closely to
orthopedics. It is clearly aligned with musculoskeletal trauma and joint
pathology. and dominated by injury-related terms (fracture, ligament,
tendon).

This is consistent with TF-IDF but provides richer thematic grouping
rather than just keyword prominence.

### Document–topic distributions (γ) by specialty

The γ matrix tells us how much each document (i.e., transcript in this
case) draws from each topic. We join it with specialty labels to see
whether LDA-discovered topics align with the known clinical categories:

```{r lda-gamma}
lda_gamma <- tidy(emr_lda, matrix = "gamma")

lda_gamma_specialty <- lda_gamma |>
  left_join(
    filtered_data |>
      mutate(document = as.character(note_id)) |>
      select(document, medical_specialty),
    by = "document"
  )
```

```{r lda-gamma-plot}
lda_gamma_specialty |>
  group_by(medical_specialty, topic) |>
  summarise(mean_gamma = mean(gamma), .groups = "drop") |>
  mutate(topic = factor(topic, labels = paste("Topic", seq_len(k_topics)))) |>
  ggplot(aes(topic, mean_gamma, fill = medical_specialty)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(title = "Mean topic proportion by specialty (γ)",
       x = NULL, y = "Mean γ", fill = "Specialty")
```

**Q12 (3 marks):**

(a) Based on the γ plot, which specialty has the clearest dominant topic
    (i.e., one topic has a noticeably higher mean γ than the others for
    that specialty)? Does this align with the clinical label you
    assigned to that topic in Q11?

    Orthopedic has the clearest dominant topic 3. Yes, it matches.

(b) Uncomment and complete the code below to compute, for each
    specialty, the single topic with the highest mean γ. Replace each
    `___` with the correct column name or expression.

```{r q12b}
lda_gamma_specialty |>
  group_by(medical_specialty, topic) |>
  summarise(mean_gamma = mean(gamma), .groups = "drop") |>
  group_by(medical_specialty) |>
  slice_max(mean_gamma, n = 1)
```

(c) Sun et al. (2024) ran LDA on social work notes *without knowing in
    advance* what topics would emerge. How does their situation differ
    from this exercise, where we set `k = 3` to match three known
    specialties? What does this suggest about how to choose the number
    of topics *k* in a real exploratory analysis?

    They used LDA in a fully exploratory setting, so the model was free
    to discover whatever number and type of topics best represented the
    data. In contrast, this exercise sets k = 3 specifically because we
    already know there are three specialties, which makes the topic
    structure partially supervised and forces the model to align with a
    known external grouping rather than discover it naturally. This
    difference suggests that in real exploratory analysis, k should not
    be chosen based on external labels, but instead tuned using a
    combination of interpretability, model fit (e.g., perplexity or
    coherence scores), and stability across runs. The goal is to find a
    value of k that produces coherent, distinct, and meaningful topics,
    rather than one that simply matches known categories or imposes
    structure on the data.

------------------------------------------------------------------------

## LLMs for Clinical Text Analysis

Finally, let's address the LLM in the room. Chung et al. (2024) showed
that GPT-4 Turbo, with no task-specific training, could classify
perioperative outcomes from pre-operative notes using carefully designed
prompts (zero-shot, few-shot, chain-of-thought).

Let's experiment similarly to classify a single note and then compare
its reasoning with the LDA topic model's assignment. You can use
whichever LLM you are comfortable using (Dalhousie students have access
to Microsoft Copilot via Office 365 (copilot.microsoft.com) with some
degree of data protection/agreements).

### Sample note

```{r sample-note, cache=FALSE}
# Sample from notes that appear in the topic model
set.seed(42)
dtm_note_ids <- as.integer(rownames(emr_dtm))

q10_note <- filtered_data |>
  filter(note_id %in% dtm_note_ids) |>
  slice_sample(n = 1)

cat("=== True specialty:", q10_note$medical_specialty, "===\n\n")
cat("Description:", q10_note$description, "\n\n")
cat("--- Transcription (first 1200 characters) ---\n")
cat(str_trunc(q10_note$transcription, 1200))
```

LDA topic assignment for this note (γ values, highest first):

```{r sample-note-gamma, cache=FALSE}
lda_gamma_specialty |>
  filter(document == as.character(q10_note$note_id)) |>
  mutate(topic = paste("Topic", topic)) |>
  arrange(desc(gamma))
```

**Q13 (3 marks):** Copy the transcription above into Microsoft Copilot
and use the following prompt:

> *"I will give you a medical transcription. Classify it as exactly one
> of: Orthopedic, Radiology, or Neurology. Think step by step and
> explain your reasoning before giving your final answer. Transcription:
> [paste text here]"*

(a) What specialty did the LLM predict, and was it correct? Does it
    match the topic with the highest γ from the LDA model above?

    Orthopedic. It is correct and matches the topic above.

(b) What specific features of the text did the LLM cite to justify its
    classification?

    Conditions mentioned and procedures performed

(c) Compare LLM's explanation-based classification with LDA's
    unsupervised topic assignment. Which approach would you prefer for
    classifying a large batch of new clinical notes, and why?

    I would generally prefer an LLM-based approach, because it is more
    accurate, interpretable at the document level, and better at
    handling complex clinical language, abbreviations, and
    context-dependent meaning.

------------------------------------------------------------------------

## Tying back to the literature

**Q14 (1 mark):** Sun et al. (2024) applied LDA topic modelling to
clinical social work notes to explore SDoH factors. Name two of the
SDoH-related topics they identified and state the name of the metric
they used to evaluate topic quality (hint: it measures the coherence of
the top words within a topic).

Mental health, family

**Q15 (1 mark):** Chung et al. (2024) tested GPT-4 Turbo on both
classification and numerical prediction tasks. Name one task where the
LLM performed well (give the F1 score) and one where it failed to beat a
simple baseline (give the MAE and the baseline value for comparison).

LLM performed well on ASA physical status (F1 score: 0.55). It failed to
beat the baseline on hospital duration prediction. 4.55 days MAE vs 5.4
days for the baseline.

------------------------------------------------------------------------

## Reproducibility

```{r session-info, cache=FALSE}
sessioninfo::session_info()
```

------------------------------------------------------------------------

*Dataset: MTSamples via `clinspacy` (mtsamples.com) with material drawn
from Silge & Robinson,* Text Mining with R *(tidytextmining.com)*
